---
title: "Power and Inequality: a Study on Democracy and the Economic Development of Nations"
author: "Isabel Dean"
description: "GOV50 Final Project"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction
Is undemocratic government linked to weaker economic development of a country? This study aims to investigate the correlation between the concentration of presidential power and a country's economic performance, as measured by the Gini Index. My main hypothesis proposes that there is a positive correlation between greater presidential power and a higher Gini Index, implying that nations with a more pronounced power imbalance tend to exhibit increased income inequality.

Throughout history, dictatorships and authoritarian regimes have experienced severe economic hardships. By using presidential power as a lens to examine the illiberalism of a country and the Gini Index as a metric for economic development, I seek to examine the intrinsic links between political governance and economic outcomes. My null hypothesis posits that there is no correlation between greater presidential power and economic development.

Understanding this relationship is vital due to its potential implications for governance, social equity, and economic prosperity. Dictatorships and authoritarian regimes have historically grappled with economic challenges; by exploring the connection between concentrated political power and economic development we can discover the true impact of democracy on the economic well being of a country. To truly understand the links between democracy and economy, it is significant to explore other factors of economic growth, GNP per capita, GDP, trade and more. Further, we must consider that measures of democracy are difficult to observe- presidential power is one lens through which I have considered the spread and balance of liberalism across countries. 

## Data
My data is taken from two sources: (1) [Presidential Power Scores](https://presidential-power.net/?page_id=2151) and (2) [World Income Inequality Database - WIID4](https://www.wider.unu.edu/sites/default/files/WIID/PDF/WIID4%20User%20Guide.pdf). The explanatory variables of interest are Country, Year, Presidential Power, Gini Index and Continent. From Continent, I then selected from my data set countries in Latin America, so as to maintain a similar standard across countries with similar countries and external factors- this way I could avoid as much confoundment as possible. My key dependent variable is income inequality, which is the Gini Index, and my independent variable is Presidential Power. These have been measured through a survey. 

```{r}
#Reading, Merging and Cleaning the Data set! 

library(tidyverse)
library(ggplot2)
pres_powers <- read.csv("PrespowData_FINAL.csv")

inequality <- read.csv("inequality.csv")

final_dataset <- pres_powers |> 
  left_join(inequality, by = c("country", "year")) |>
  select(country, year, prespow1, gini_reported, regioncode) |>
  drop_na() |>
  filter(regioncode == "Americas")

final_dataset |>
 knitr::kable(col.names = c("Country", "Year", "Presidential Power", "Gini Index", "Continent"))
```

Through my cross-sectional design, I hope to analyse the correlation between Presidential Power and higher Gini Coefficients across Latin American countries in the same year, which I have achieved through merging my data sets, joining through country and year. The variables I am measuring are continuous- Presidential Power is measured on a scale of 0-1, with 1 being total Presidential power, whilst the Gini Coefficient is measured on a scale of 0-100, with 100 being complete income inequality. 

```{r}
#A Plot Summarizing Income Inequality Across Countries 

final_dataset |>
  ggplot(mapping = aes(x = year, color = country)) +
  geom_point(mapping = aes(y = gini_reported)) +
  labs(x = "Year", 
       y = "Gini Index",
       title = "Latin American Income Inequality",
       subtitle = "(Nations have been measured at different points from 1960-2001)")
```

As the scatter plot shows, while we cannot see most of the individual progression of countries over time, income inequality is higher toward the last 15 years of the 20th century among Latin American countries. There is one notable outlier, being that Colombia in 1960 had a Gini Index of 59.20. This can likely be explained by contingent factors that won't be explored further in this project. From this data set, the lowest Gini Index was Panama in 1972	at 34.80, and the highest was Brazil in 1988 at 61.43. For context, ["Economies with Gini values above (50) are considered very unequal... A Gini (Index) below (30) is considered low... (and) In most countries the Gini coefficient lies between (30 and (50). In many advanced economies, there has been an increase in income inequality since the 1980s."](https://wol.iza.org/articles/measuring-income-inequality/long#:~:text=A%20Gini%20coefficient%20below%200.3,since%20the%201980s%20%5B10%5D.) 

## Results
```{r}
#Scatterplot of Main Relationship of Interest

main_regression <- final_dataset |>
  ggplot(mapping = aes(y = gini_reported)) +
  geom_point(mapping = aes(x = prespow1)) +
  geom_smooth(mapping = aes(x = prespow1), method = "lm", se = FALSE) +
  labs(x = "Presidential Power",
       y = "Gini Index",
       title = "The Relationship between Presidential Power and the Gini Index")

main_regression
```

```{r}
#R Squared and P Value

library(broom)

project_regression <- lm(data = final_dataset, gini_reported ~ prespow1)

glance(project_regression) |>
  select(r.squared, p.value) |>
  knitr::kable(digits = 3, col.names = c("R Squared", "P Value"))
```

```{r}
#Regression 

project_regression <- lm(data = final_dataset, gini_reported ~ prespow1)

var_labels <- c(
  "(Intercept)" = "Intercept",
  "prespow1" = "Presidential Power"
  ) 

modelsummary::modelsummary(project_regression,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           coef_map = var_labels,
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))

```

My regression scatterplot and line of best fit of Presidential Power plotted against the Gini Index reveal that there is a weak positive relationship between the two. The slope 0.125 illustrates that as as Presidential Power increases by 1 unit, income inequality rises on average by 0.125 units on the Gini Index. With an R Squared value less than 0.3, this correlation is weak, and above 0, positive, however this essentially shows that when Presidential Power is higher, income inequality tends to be somewhat higher as well. 

Because of this weak, positive correlation, this evidence alone is not enough to support the hypothesis that a greater Presidential Power has a causal relationship with income inequality. However, with a P-value that is below 0.05 at 0.02, these results are statistically significant. At the 0.05 significance level, but with a weak correlation, it is a model that could lead us in the right direction, however does not show the full picture. 

## Conclusion
Overall, my null hypothesis was disproved: there was a positive correlation between higher Presidential Power and higher income inequality. Despite this relationship being statistically significant, with a low coefficient, this relationship is a weak one, which does not provide enough evidence to attribute a causal relationship. In this way, my main hypothesis can only be weakly supported, and requires further investigation on other fronts to prove. There were several limitations upon my analysis. First, a better model could have been used to assess the relationship between Presidential Power and Income Inequality- namely, a Differences in Differences Design would be able to show how changes in the levels of Presidential Power affect changes in Income Inequality, rather than measuring their relationship at a specific point in time. Further, there are multiple confounders that should be studied, such as economic growth, transnational economic influences, policy and government style outside of presidency, other democratic measures such as Freedom House and Polity IV. Through examining these, we may find stronger links between economic development and the state of democracy in nations. To improve my analysis, I would analyse countries over time, investigating how the increase of Presidential Power over a time period correlates with the increase in income inequality. I would then compare these differences across countries to examine whether there is a similar pattern across nations. 
