---
title: "Power and Inequality: "
author: "Isabel Dean"
description: "My Final project"
output:
  distill::distill_article:
    self_contained: false
---

## Introduction

Is undemocratic government linked to weaker economic development of a country? This study aims to investigate the correlation between the concentration of presidential power and a country's economic performance, as measured by the Gini coefficient. My main hypothesis proposes that there is a positive correlation between greater presidential power and higher Gini coefficients, implying that nations with a more pronounced power imbalance tend to exhibit increased income inequality.

Throughout history, dictatorships and authoritarian regimes have experienced severe economic hardships. By using presidential power as a lens to examine the illiberalism of a country and the Gini coefficient as a metric for economic development, I seek to examine the intrinsic links between political governance and economic outcomes. My null hypothesis suggests that there is no correlation between greater presidential power and economic development.

Understanding this relationship is vital due to its potential implications for governance, social equity, and economic prosperity. Dictatorships and authoritarian regimes have historically grappled with economic challenges; by exploring the connection between concentrated political power and economic development we can discover the true impact of democracy on the economic well being of a country. To truly understand the links between democracy and economy, it is significant to explore other factors of economic growth, GNP per capita, GDP, trade and more. Further, we must consider that measures of democracy are difficult to observe- presidential power is one lens through which I have considered the spread and balance of liberalism across countries. 


## Data
(2) a data section that briefly describes the data source, describes how the key dependent and independent variables are measured (e.g., a survey, statistical model, or expert coding), and also produces a plot that summarizes the dependent variable; 

For the data section, you should 

- note if your research design is cross-sectional (most projects will be of this type) or one of the other designs we discussed (randomized experiment, before-and-after, differences-in-differences). CR4OSS. SECTIONINAL
- Data section: (3pts)
    - 2-3 paragraphs
        - INDEPENDENT- PRES POWER
        - DEPENDENDT- INCOME INEQUALITY
        - survey
    - plot visualizing main outcome of interest


## Results
(3) a results section that contains a scatterplot, barplot, or boxplot of the main relationship of interest and output for the main regression of interest; and 

- Results section: (8pts)
    - plot of main analysis
    - regression output
    - 2-3 paragraphs of description and interpretation of the plots and regression (including interpreting the main coefficient of interest and describing if it is statistically significant and if we should interpret it causally). This section could be longer if you choose to include additional analyses.
    
For the results section, you should interpret (in plain English) the main coefficient of interest in your regression. You should also comment on the statistical significance of the estimated coefficient and whether or not you believe the coefficient to represent a causal effect.

## Conclusion
(4) a brief (one paragraph) concluding section that summarizes your results, assesses the extent to which you find support for your hypothesis, describes limitations of your analysis and threats to inference, and states how your analysis could be improved (e.g., improved data that would be useful to collect).
- Conclusion section: 1 paragraph (2pts)
    - (i) summarizing results and assessing the extent to which you find support for your hypothesis;
    - (ii) describing limitations of the analysis and threats to inference (missing data, confounding, etc), and stating how you could improve your analysis if you had more time/money.



```{r}
library(tidyverse)
pres_powers <- read.csv("PrespowData_FINAL.csv")

inequality <- read.csv("inequality.csv")

pres_powers

final_dataset <- pres_powers |> left_join(
  inequality, by = c("country", "year")
) |>
  select(country, year, prespow1, gini_reported, regioncode) |>
  drop_na() |>
  filter(regioncode == "Americas")

final_dataset |>
 knitr::kable(col.names = c("Country", "Year", "Presidential Power", "Gini Coefficient", "Continent"))

```

```{r}
#(2) a data section that briefly describes the data source, describes how the key dependent and independent variables are measured (e.g., a survey, statistical model, or expert coding), and also produces a plot that summarizes the dependent variable; 
final_dataset |>
  ggplot(mapping = aes(x = year, color = country)) +
  geom_point(mapping = aes(y = gini_reported))

final_dataset
```

```{r}
graph_1 <- final_dataset |>
  ggplot(mapping = aes(y = gini_reported)) +
  geom_point(mapping = aes(x = prespow1)) +
  geom_smooth(mapping = aes(x = prespow1), method = "lm", se = FALSE)


graph_1
```
```{r}
project_regression <- lm(data = final_dataset, gini_reported ~ prespow1)
```

```{r}
library(broom)

glance(project_regression) |>
  select(r.squared, p.value) |>
  knitr::kable(col.names = c("R Squared", "P Value"))
```

```{r}
project_regression <- lm(data = final_dataset, gini_reported ~ prespow1)

modelsummary::modelsummary(project_regression,
                           statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),
                           gof_map = c("nobs", "r.squared", "adj.r.squared"))
```


