---
title: "Gov 50 Final Project"
author: "Isabel Dean"
description: "My Final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project Thoughts

I am interested in exploring data related to finding links between the reasons behind not voting, different demographics and participation in elections. Coming from a country that mandates elections, it is interesting to me that countries like America seem to have surprisingly low election participation. I have heard many reasons tossed around, such as access to polling, disparities between different demographics, education and more. Thus, I would like to analyse the different reasons and evident results of voters to discover the main reasons as to why people don't vote.  

## Project Milestone 2 Proposal

What is the main reason that people don't vote? In America's most recent presidential election, only 66% of eligible voters participated in the 2020 election, which was a record high. It is necessary to study why eligible voters are failing to participate in elections, so that we can use this data to locate the measures we can take to increase universal suffrage and voter turnout. Likely, the largest factors that lead people to not vote in elections will be barriers to voting such as waiting in line, taking time off work or missing the voter registration deadline. These factors itself are linked to income group, race, etc which can cause confoundment issues and should be accounted for. My hypothesis from these considerations will be that the largest reason as to why people don't vote will be at the highest intersection between a particular demographic and access to voting, which seems likely to be that lower income groups are least likely to vote due to inaccessibility in participating in elections. My data is taken from the Why Many Americans Don't Vote data set done by Ipsos for FiveThirtyEight. The explanatory variables of interest are Questions 5, 8, 11, 16-18, 20-22, 25, 28, 29, which cover political opinions, trust or distrust of social and political institutions, accessibility to vote, whether or not respondents are registered to vote and reasons to or not to vote. These variables are coded 1 to a certain number such as 1-4, or 1-8, that indicates different discrete answers to the questions- thus I can measure the answers that are highest chosen by voters. I will analyse these answers through the categorical variables of respondents including voter_category(consistent, somewhat or never voters), ppage(age of voter), gender(gender of voter), educ(highest education attained), race(race of voter), income_cat(household income category of voter). This categorical data will be used to measure trends among different classifications of respondents. My outcome variable will be the failure to vote in elections. My hypothesis would be proved if there was a positive, significant coefficient when I run my regression between low income groups and reported inaccessibility to voting, as well as a positive, significant coefficient representing the link between this respondent group and a failure to vote. If there is a negative or weak correlation that is not statistically significant, then my hypothesis would be disproved. From analysing multiple variables, I may be led to a new conclusion about the demographic and reasons why most people don't vote. 

## Project Milestone 3 Data Visualisation

I have shown the proportion of non-voters by income group with my column graph. I have compared respondents from different income groups against the proportion of each group that indicated they planned not to vote in the 2020 November Presidential election. Thus, I have shown that as income increases, the likelihood a respondent plans not to vote decreases. The highest proportion of those who planned not to vote was 12.8% of respondents who made less than 40,000 a year. The lowest percentage was at 4.5% belonging to respondents who made 125,000 or more per year. 

```{r}
library(tidyverse)
library(readr)
nonvoters <- read_csv("nonvoters_data.csv")
nonvoters
```

```{r}
# Load necessary libraries
library(ggplot2)
library(tidyverse)
library(readr)
nonvoters <- read_csv("nonvoters_data.csv")

nonvoters1 <- nonvoters |>
  group_by(income_cat) |>
  summarise(Proportion_Not_Voting = mean(Q21 == 2))

visualisation_1 <- nonvoters1 |>
  mutate(income_cat = fct_relevel(income_cat, "Less than $40k", "$40-75k", "$75-125k", "$125k or more")) |>
  ggplot(aes(x = income_cat, y = Proportion_Not_Voting)) +
  geom_col() +
  labs(title = "Proportion of Non-Voters by Income Group",
       x = "Income Group",
       y = "Proportion Planned Not to Vote")

visualisation_1
```

## Project Milestone 4 Data Analysis
From the first analysis done in Milestone 3, with 'Proportion of Non-Voters by Income Group', I observed that as income decreases, the likelihood a respondent plans not to vote increases. The highest proportion of those who planned not to vote was 12.8% of respondents who made less than 40,000 a year. With my next analysis, 'Non-Voter Beliefs about Voting Split by Income Group', I aimed to see whether beliefs about voting differed between income groups. Using the variable Q29_3, I found that of those who do not vote, a much higher proportion agree with the statement "No matter who wins, nothing will change for people like me" and find this an important reason as to why they don't vote. However, of those who resonate with this reason and those who don't, the proportions of these groups in terms of income class do not differ greatly. This suggests that attitudes about belonging and change affect those of all income groups- a regression analysis would further reveal the correlation, or lack of correlation between income group and attitudes towards change based on identity groups. My next graph, 'Respondents of different Income Groups Voting Choices' reveals the different proportions of voting choice groups. The column graph illustrates that those in the income group 'Less than 40k' take up higher proportions of those who don't, or sporadically vote than in always voting. This is in contrast to other income groups, which take up similar proportions of always, rarely/never and sporadic groups, demonstrating the disproportionate difference when it comes to the lowest income group not voting. This analysis shows that investigating reasons as to why people don't vote will likely have a correlation to lower income groups, specifically those who make 'Less than 40k' a year. 

```{r}
library(ggplot2)

nonvoters <- nonvoters |>
  mutate(reason_disillusioned = if_else(Q29_3 == 1, "No matter who wins, nothing will change for people like me", "Not an Important Factor"), na.rm = TRUE) 

graph_2 <- nonvoters|>
  filter(reason_disillusioned != 'NA') |>
  ggplot(mapping = aes(x = reason_disillusioned, fill = income_cat)) +
  geom_bar() +
  labs(x = "Non-Voters",
       y = "Number of Non-Voters",
       title = "Non-Voter Beliefs about Voting Split by Income Group" )

graph_2
```

```{r}
graph_3 <- nonvoters |>
  ggplot(mapping = aes(x = voter_category, fill = income_cat)) +
  geom_bar() +
  labs(y = "Number of Respondents",
       x = "Voting Choice",
       title = "Respondents of different Income Groups' Voting Choices")

graph_3
```

